---
title: "WHOOP Biometrics · Personal Dashboard"
subtitle: "13 months of physiological data — March 2023 to April 2024"
---

```{r setup}
source("R/load_data.R")
library(patchwork)
library(ggridges)
```

This site explores a year-plus of personal data from the WHOOP wearable, applying exploratory analysis techniques inspired by the *Functional Data Analysis with R* textbook. Each daily measurement is treated as a sample from an underlying physiological process — the goal is to visualize longitudinal patterns, identify structure, and surface insights that the WHOOP app alone doesn't show.

---

## At a glance

```{r metric-cards, echo=FALSE}
htmltools::HTML(glue::glue('
<div class="metric-grid">
  <div class="metric-card blue">
    <div class="label">Days tracked</div>
    <div class="value">{n_days}</div>
    <div class="sub">{date_range}</div>
  </div>
  <div class="metric-card green">
    <div class="label">Avg recovery</div>
    <div class="value">{avg_recovery}%</div>
    <div class="sub">on the WHOOP 0–100 scale</div>
  </div>
  <div class="metric-card blue">
    <div class="label">Avg HRV</div>
    <div class="value">{avg_hrv} ms</div>
    <div class="sub">resting baseline</div>
  </div>
  <div class="metric-card red">
    <div class="label">Avg RHR</div>
    <div class="value">{avg_rhr} bpm</div>
    <div class="sub">resting heart rate</div>
  </div>
  <div class="metric-card amber">
    <div class="label">Total workouts</div>
    <div class="value">{n_workouts}</div>
    <div class="sub">{n_swims} swimming sessions</div>
  </div>
  <div class="metric-card green">
    <div class="label">Peak HRV</div>
    <div class="value">{peak_hrv} ms</div>
    <div class="sub">{format(peak_hrv_date, "%b %d, %Y")}</div>
  </div>
</div>
'))
```

---

## Recovery over time

The plot below shows daily recovery scores with a 7-day rolling mean smoothing out day-to-day noise — analogous to the functional mean estimators in Chapter 1 of the FDA textbook.

```{r recovery-overview, fig.cap="Daily recovery (points) with 7-day rolling mean (line). Color encodes WHOOP recovery zones: green ≥ 67%, yellow 34–66%, red < 34%."}
ggplot(cycles, aes(x = date)) +
  geom_point(aes(y = recovery, color = rec_zone), alpha = 0.45, size = 1.2, shape = 16) +
  geom_line(aes(y = recovery_7), color = "#111111", linewidth = 0.9, na.rm = TRUE) +
  scale_color_manual(
    values = c(Green = "#1D9E75", Yellow = "#BA7517", Red = "#E24B4A"),
    name = "Zone"
  ) +
  scale_x_date(date_breaks = "2 months", date_labels = "%b '%y") +
  scale_y_continuous(limits = c(0, 100), breaks = c(0, 33, 67, 100)) +
  geom_hline(yintercept = c(34, 67), linetype = "dashed", color = "grey70", linewidth = 0.4) +
  labs(
    title    = "Daily recovery score",
    subtitle = "7-day rolling mean overlaid in black",
    x = NULL, y = "Recovery (%)"
  ) +
  theme_whoop()
```

---

## Recovery zone distribution

```{r zone-dist, fig.height=3.5, fig.cap="Monthly proportion of days in each recovery zone, showing how readiness varied across the year."}
cycles |>
  filter(!is.na(rec_zone)) |>
  count(month, rec_zone) |>
  group_by(month) |>
  mutate(pct = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(x = month, y = pct, fill = rec_zone)) +
  geom_col(width = 25) +
  scale_fill_manual(
    values = c(Green = "#1D9E75", Yellow = "#BA7517", Red = "#E24B4A"),
    name = "Zone"
  ) +
  scale_x_date(date_breaks = "2 months", date_labels = "%b '%y") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title    = "Monthly recovery zone composition",
    subtitle = "Stack shows proportion of Green / Yellow / Red days per month",
    x = NULL, y = "Proportion of days"
  ) +
  theme_whoop()
```

---

## What's in each section

| Page | Content |
|---|---|
| [Recovery & HRV](recovery.qmd) | HRV trajectories, RHR trends, strain relationship |
| [Sleep](sleep.qmd) | Stage composition, sleep debt, consistency |
| [Workouts](workouts.qmd) | Swim analysis, timing patterns, HR zones |
| [Methods](methods.qmd) | Data provenance, smoothing approach, notes |
