---
title: "Methods & Data Notes"
subtitle: "Data provenance, processing decisions, and analytical approach"
---

```{r setup}
source("R/load_data.R")
library(dplyr)
```

## Data source

All data come from the **WHOOP 4.0** wearable worn continuously from `r format(min(cycles$date), "%B %d, %Y")` through `r format(max(cycles$date), "%B %d, %Y")`. Three CSV exports were used:

| File | Content | Rows |
|---|---|---|
| `physiological_cycles.csv` | Daily recovery, HRV, RHR, strain, and embedded sleep summary | `r nrow(cycles_raw)` |
| `sleeps.csv` | Individual sleep sessions including naps, with stage breakdown | `r nrow(sleeps_raw)` |
| `workouts.csv` | Individual workout sessions with HR zones and strain | `r nrow(workouts_raw)` |

Data were exported directly from the WHOOP app under *Account → My Data → Export*.

---

## Processing decisions

**Physiological cycles.** Rows with missing `Recovery score %` were dropped (`r sum(is.na(cycles_raw[["Recovery score %"]]))` rows removed). Rows where RHR > 80 bpm appear in the raw data from a known multi-day anomaly period (late July 2023) and are retained in the dataset but filtered out of correlation plots where noted.

**Sleeps.** The `sleeps.csv` file includes both overnight sleeps and naps (flagged via the `Nap` column). Naps are excluded from stage composition and sleep debt analyses but included in session counts.

**Workouts.** Activity names were consolidated from `r n_distinct(workouts_raw[["Activity name"]])` distinct labels into 5 groups (Swimming, Lifting, Running, Basketball, Other). The `Strength Trainer` label was merged into Lifting.

**Rolling means.** All 7-day rolling averages use `zoo::rollmean()` with `align = "right"` and `na.rm = TRUE`, meaning the value at day *t* reflects the mean of the current and previous 6 days with available data. The 28-day rolling means on the sleep page use the same approach.

---

## Analytical inspiration

This project is modeled on the exploratory visualization approach in:

> Goldsmith, J., Scheipl, F., Huang, L., Wrobel, J., Di, C., Gellar, J., ... & Crainiceanu, C. (2024). *Functional Data Analysis with R*. Chapman & Hall / CRC.

Specific techniques borrowed:
- **Spaghetti + mean overlay** (Chapter 1 CD4 example) → Recovery & HRV page
- **Stacked composition charts** (COVID mortality page) → Sleep stage composition
- **Circadian timing plots** (NHANES MIMS page) → Workout timing analysis
- **Scatter with smooth** (CD4 individual trajectory plots) → Strain vs. recovery scatter

---

## Rendering this site

This is a **Quarto website** rendered with `quarto render` and deployed via GitHub Pages from the `docs/` folder.

**R packages required:**

```r
install.packages(c(
  "tidyverse",
  "lubridate",
  "zoo",
  "patchwork",
  "ggridges",
  "glue",
  "htmltools",
  "scales"
))
```

**To render locally:**

```bash
# From the project root
quarto render

# Or preview with live reload
quarto preview
```

---

## Known limitations

- WHOOP's HRV measurement uses a 5-minute morning calibration window, which can be sensitive to movement or breathing irregularities — single-day outliers should not be over-interpreted.
- The sensor was removed during the late July 2023 anomaly period (RHR showing 120 bpm is a known artifact when the band is off but still recording).
- Sleep stage classification is proprietary and not validated against polysomnography.
- Workout start times are rounded to the hour in the export format.

---

## Session info

```{r session-info}
sessionInfo()
```
