---
title: "Recovery & HRV"
subtitle: "Heart rate variability trajectories and autonomic nervous system readiness"
---

```{r setup}
source("R/load_data.R")
library(patchwork)
```

HRV — the millisecond variation between heartbeats — is the closest non-invasive proxy for autonomic nervous system readiness. Higher HRV generally signals better recovery; lower HRV reflects accumulated stress, illness, or overtraining. WHOOP measures it each morning during a brief calibration window.

---

## HRV over time

```{r hrv-timeline, fig.height=5.5, fig.cap="Daily HRV (points) and 7-day rolling mean (line). Colored bands mark the interquartile range across the full period. Notable: the July 2023 RHR/HRV anomaly is visible as several days of very low HRV coinciding with an RHR spike to 120 bpm — likely illness or injury."}

# IQR reference band
hrv_q <- quantile(cycles$hrv, probs = c(0.25, 0.75), na.rm = TRUE)

ggplot(cycles, aes(x = date)) +
  annotate("rect",
    xmin = min(cycles$date), xmax = max(cycles$date),
    ymin = hrv_q[1], ymax = hrv_q[2],
    fill = "#3266ad", alpha = 0.07
  ) +
  geom_point(aes(y = hrv, color = rec_zone), alpha = 0.4, size = 1.3, shape = 16) +
  geom_line(aes(y = hrv_7), color = "#3266ad", linewidth = 1, na.rm = TRUE) +
  geom_hline(yintercept = avg_hrv, linetype = "dotted", color = "grey50") +
  scale_color_manual(
    values = c(Green = "#1D9E75", Yellow = "#BA7517", Red = "#E24B4A"),
    name = "Recovery zone"
  ) +
  scale_x_date(date_breaks = "2 months", date_labels = "%b '%y") +
  annotate("text", x = min(cycles$date) + 5, y = avg_hrv + 4,
           label = paste0("Mean = ", avg_hrv, " ms"), size = 3, color = "grey50") +
  labs(
    title    = "Heart rate variability (HRV)",
    subtitle = "Blue band = interquartile range · Dotted line = overall mean",
    x = NULL, y = "HRV (ms)"
  ) +
  theme_whoop()
```

---

## HRV vs. Resting Heart Rate

RHR and HRV should move in opposite directions under healthy adaptation — high HRV paired with low RHR is the ideal combination. The scatter below shows this inverse relationship in your data.

```{r hrv-rhr-scatter, fig.height=4.5, fig.cap="Each point is one day. The negative correlation between HRV and RHR is clearly visible; outliers in the top-right corner (high HRV *and* high RHR) typically coincide with illness events. Points filtered to RHR < 80 to exclude sensor artifacts."}
cycles |>
  filter(!is.na(hrv), !is.na(rhr), rhr < 80) |>
  ggplot(aes(x = rhr, y = hrv, color = rec_zone)) +
  geom_point(alpha = 0.5, size = 1.8, shape = 16) +
  geom_smooth(method = "lm", se = TRUE, color = "#111", linewidth = 0.8,
              fill = "grey85", alpha = 0.5) +
  scale_color_manual(
    values = c(Green = "#1D9E75", Yellow = "#BA7517", Red = "#E24B4A"),
    name = "Recovery zone"
  ) +
  labs(
    title    = "HRV vs. resting heart rate",
    subtitle = "Linear trend with 95% confidence band",
    x = "Resting heart rate (bpm)", y = "HRV (ms)"
  ) +
  theme_whoop()
```

```{r hrv-rhr-corr}
cor_val <- cor(cycles$hrv, cycles$rhr, use = "complete.obs") |> round(3)
```

> The Pearson correlation between daily HRV and RHR is **r = `r cor_val`**, confirming the expected inverse relationship.

---

## Rolling means: HRV, RHR, and Strain

This three-panel view (modeled after the NHANES multi-panel plots in the FDA textbook) shows 7-day rolling means for all three key recovery metrics simultaneously, making it easier to spot periods where all three converged or diverged.

```{r three-panel, fig.height=7, fig.cap="7-day rolling averages for HRV, RHR, and day strain. Shaded regions highlight the two major illness/anomaly events (early April 2023 and late July 2023)."}

illness_bands <- tibble(
  xmin = as_date(c("2023-04-03", "2023-07-26")),
  xmax = as_date(c("2023-04-15", "2023-07-31"))
)

make_band <- function(p) {
  p + geom_rect(
    data = illness_bands,
    aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf),
    fill = "#E24B4A", alpha = 0.08, inherit.aes = FALSE
  ) +
  scale_x_date(date_breaks = "2 months", date_labels = "%b '%y") +
  theme_whoop() +
  labs(x = NULL)
}

p1 <- ggplot(cycles, aes(x = date, y = hrv_7)) +
  geom_line(color = "#3266ad", linewidth = 0.9, na.rm = TRUE) +
  labs(title = "HRV — 7-day rolling mean", y = "ms")

p2 <- ggplot(cycles, aes(x = date, y = rhr_7)) +
  geom_line(color = "#E24B4A", linewidth = 0.9, na.rm = TRUE) +
  labs(title = "Resting heart rate — 7-day rolling mean", y = "bpm")

p3 <- ggplot(cycles, aes(x = date, y = strain_7)) +
  geom_line(color = "#BA7517", linewidth = 0.9, na.rm = TRUE) +
  labs(title = "Day strain — 7-day rolling mean", y = "strain score")

make_band(p1) / make_band(p2) / make_band(p3)
```

::: {.callout-note}
**Red shaded regions** mark two anomalous periods: early April 2023 (recovery crashed to 1% on consecutive days, HRV dropped sharply) and late July 2023 (RHR spiked to 120 bpm for several days — almost certainly illness or an injury event).
:::

---

## Day-of-week patterns

```{r dow-box, fig.height=4, fig.cap="Distribution of HRV by day of week. Sunday typically shows the highest HRV (post-rest), while mid-week shows more variability from accumulated training load."}
cycles |>
  filter(!is.na(hrv)) |>
  ggplot(aes(x = day_of_week, y = hrv, fill = day_of_week)) +
  geom_boxplot(alpha = 0.7, outlier.size = 1, outlier.alpha = 0.4, width = 0.6) +
  geom_jitter(width = 0.15, alpha = 0.2, size = 0.8) +
  scale_fill_manual(values = rep(c("#3266ad","#1D9E75","#BA7517","#E24B4A",
                                    "#7F77DD","#888780","#3266ad"), 1)) +
  labs(
    title    = "HRV by day of week",
    subtitle = "Boxplot with jittered raw values overlaid",
    x = NULL, y = "HRV (ms)"
  ) +
  theme_whoop() +
  theme(legend.position = "none")
```

---

## Recovery score distribution

```{r rec-hist, fig.height=3.5}
cycles |>
  ggplot(aes(x = recovery, fill = rec_zone)) +
  geom_histogram(binwidth = 5, color = "white", linewidth = 0.3) +
  geom_vline(xintercept = avg_recovery, linetype = "dashed", color = "#111") +
  scale_fill_manual(
    values = c(Green = "#1D9E75", Yellow = "#BA7517", Red = "#E24B4A"),
    name = "Zone"
  ) +
  annotate("text", x = avg_recovery + 2, y = 30, label = paste0("Mean = ", avg_recovery, "%"),
           size = 3.2, hjust = 0, color = "#111") +
  labs(
    title = "Recovery score distribution",
    subtitle = "Across all tracked days",
    x = "Recovery (%)", y = "Days"
  ) +
  theme_whoop()
```
